Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-11 Manuscript/manuscript.tex

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability in the classical MD simulation
models has been considered higly relevant issue since the early days of
lipid bilayer simulations. Some pioneering simulation studies scaled
the partial charges of atoms by one half to effectively include 
polarizability \cite{jonsson86,egberts94}. Also, approaches that
explicitly include polarizability has been introduced, but
this has turned out to be practically complicated \cite{chowdhary13}.
In this work we effectively include electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
In contrast to those studies, however, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

According to ECC, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for partial charges in molecules,
i.e. lipids in our case. Unlike the total charge of ions, atomic partial charges in
classical MD simulations are not experimental observables.
Instead they are based on theoretical calculations and the
derived values depend, for example, on the conformational
sampling of the molecules and description of solvent 
in the calculations \cite{RESP_paper,??}. 
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~\cite{RESP_paper}),
Depending to the details, partial charge calculations may partially
include some of the solvent electronic polarizability effects. Thus, we expect that
the correct ECC scaling factor, $f_q$, for molecular partial charges would lie
between $\epsilon _{el} ^{-1/2}$ (no electronic polarizability in calculation)
and $1$ (all electronic polarizability in calculation).

In this work, we empirically sample this parameter space
to find a atomistic MD simulation force field parameters for lipids that
accurately describes the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
when compared with NMR data \cite{akutsu81,altenbach84}. This data can used to accurately asses
ion binding in PC bilayers, as discussed in Ref, \citenum{catte16} and in section \ref{section:electrometer}.
We chose the Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and headgroup
parameters are required in all available lipid models~\cite{botan15}.

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters of the same atoms for which charges were scaled
with a scaling factor of $f_\sigma = 0.89$, the area per molecule value was
againg in agreement with experimental value (see Table~\ref{tab:apls}).

@@

@@

\subsection{POPC membrane and electrometer concept with cationic surfactant}

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2_surf.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration and
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) mole fraction
    from simulations and experiments (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, surfactant \cite{scherer89}).
  }
  \todo{This will be replaced with new figure with surfactants and P-N vector analysis when it is done.}
\end{figure*}

@@

Changes of headgroup order parameter as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}).
The head group order parameter response to the bound cation concentration
is approximately linear at least up to $\sim$0.3 mole fraction in experiments and
in both simulation models, original Lipid14 and ECClipids \todo{Yet to be shown.}.

Quantitative comparison, however, reveals that the response is overestimated in 
Lipid14 for both segments, while the ECC-lipids model gives a
good agreement for the change of $\alpha$ segment order parameter, and slightly
underestimates the $\beta$ segment order parameter change.
%Such a finding is not surprising, as the $\Delta S^\alpha/\Delta S^\beta$ ratio 
%of the response of Lipid14 model is found to be slightly larger than in 
%experiments (VALexp vs VALsim
%\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
%  SAMULI: Maybe these could be put in the same figure with the order parameter
%  response and P-N vector angle change with surfactants.
%}
%), and this property is also kept by the newly derived Lipid14-based model, ECC-lipids. 
The present comparison of head group order parameter changes 
suggests that the observed overestimated response of all models to 
increasing CaCl$_2$ concentration in \cite{catte16} 
can be explained at least in part by a high sensitivity of the head group order parameter response
to the bound charge.

\subsection{POPC membrane and electrometer concept with cation binding}

@@

@@

After achieving accurate head group order parameter response with a cationic surfactant, 
we advanced to quantifying the response to solutions with varying concentrations of salts, namely \ce{NaCl} and \ce{CaCl_2}. 
Headgroup order parameter response to increasing CaCl$_2$ concentration
(as defined in Equation~\ref{eq:conc})
from experiments, original Lipid14 model and ECC-lipids model are shown
in Fig. \ref{OrderParameterCHANGESnewMODELS}. The order parameter response
is significantly overestimated in original Lipid14 model, while results from
ECC-lipids model are in a good agreement with the experiments.
This is a significant improvement over current models,
which in general overstimate the head group order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicate
that the binding details of \ce{Ca^2+} are realistic in the ECC-lipids model
and, hence, it can be used to study lipid-ion interaction in detail.

Based on the accurate head group order parameter response of the new model, ECC-lipids, (Fig.~\ref{OrderParameterCHANGESnewMODELS})
we suggest that it also exhibits accurate ion binding affinity to zwitterionic phospholipid membranes.
This is directly related to the amount of surface excess charge 
as shown along with ion density profiles in Fig.~\ref{fig:cacl-dens}.
The density profiles from simulations with original Lipid14 \cite{dickson14} and Dang ions \cite{smith94,chang1999,dang2006} 
show a pronounced peak in the position of the phosphate moieties of POPC. 
The use of a ECC-ion model~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}
or the standard Amber ion model by \AA{qvist}~\cite{aqvist90}
along with Lipid14 does not significantly change it 
(Headgroup order parameter responses for these models in SI).
\todo{Add OP-response of Lipid14+ECC-ions into the OP-changes plot in SI}
The new ECC-lipids model with ECC-ions~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16} exhibits on the other hand smaller density in this region 
suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
It is then likely that the models that overestimate head group order parameter response~\cite{catte16} also overestimate the amount of bound cations.

>>>>

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polariization effects \cite{chowdhary13} by effectively including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
However, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

According to ECC, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a dielectric continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of force field for ions \cite{kohagen14,kohagen16,??}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and 
improves model performance against experimental scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for partial charges in molecules,
i.e, lipids in our case. Unlike the total charge of ions, atomic partial charges in
classical MD simulations are not experimental observables.
Instead they are based on theoretical calculations and the
derived values depend, for example, on the conformational
sampling of the molecules and description of solvent 
in the calculations \cite{RESP_paper,??}. 
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~\cite{RESP_paper}),
Depending to the details, partial charge calculations may partially
include some of the solvent electronic polarizability effects. Thus, we expect that
the correct ECC scaling factor, $f_q$, for molecular partial charges would lie
between $\epsilon _{el} ^{-1/2}$ (no electronic polarizability in charge calculation)
and $1$ (full electronic polarizability in charge calculation).

In this work, we empirically sample this parameter space
to find a atomistic MD simulation force field parameters for lipids that
accurately describe the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
when compared with NMR data \cite{akutsu81,altenbach84}. This data can be used to accurately asses
ion binding in PC bilayers, as discussed in Ref, \citenum{catte16} and in section \ref{section:electrometer}.
We chose the Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and headgroup
parameters are required in all available lipid models~\cite{botan15}.

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,??}.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters of the same atoms for which charges were scaled
with a factor of $f_\sigma = 0.89$, the area per molecule value was
againg in agreement with experimental value (see Table~\ref{tab:apls}).

@@

The binding affinities between different models were compared by calculating
the relative surface excess between water and ions $\Gamma_w^i$ \cite{chattorajBOOK}
\begin{equation}
  \Gamma_w^i=\Gamma_w^x-\Gamma_i^x\frac{c_w-c_w'}{c_i-c_i'},
\end{equation}
where $c_w$ and $c_i$ are concentrations in bulk water, 
$c_w'$ and $c_i'$ are concentrations inside bilayer, and
$\Gamma_w^x$ and $\Gamma_i^x$ are surface excess of water and
ions, respectively. Here consider the whole simulation box
as a interface region, thus the surface excesses
are $\Gamma_w^x=n_w/A$ and $\Gamma_i^x=n_i/A$, where
$A$ is the area of the simulation box, and $n_w$ and
$n_i$ are total number of water and ions in simulation box,
respectively. The concentrations of water and ions inside
bilayer can be expected to vanish, thus $c_w'=c_i'=0$.
Then we get the relative surface excess for lipid bilayer
simulations
\begin{equation}\label{surfexcess}
  \Gamma_w^i=\frac{n_w}{A}-\frac{n_i}{A}\frac{c_w}{c_i},
\end{equation}
where concentrations in bulk, $c_w$ and $c_i$, can be determined
from the simulation box boundary.

@@

\subsection{Response of headgroup on bound charge}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.png}
  \caption{\label{OrderParameterCHANGESsurf}
    Headgroup order parameter changes and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$)
    in PC bilayer from simulations and experiments \cite{scherer89}.
  }
  \todo{Labels are missing.} \\
  \todo{Experimental data points are missing.}\\
  \todo{Order should be the same as in other figures, i.e, $\beta$ segment on top.}
\end{figure}

@@

Changes of headgroup order parameter as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESsurf}).
The head group order parameter response to the bound cation concentration
is approximately linear at least up to $\sim$0.3 mole fraction in experiments and
in both simulation models, original Lipid14 and ECClipids \todo{Yet to be shown.}.

The slope is, however, too steep in Lipid14 model indicating that the
headgroup order parameters are too sensitive for bound charge in this model.
This has to be taken into account when interpreting the
headgroup order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.

The headgroup P-N vector angle is shown in Fig.~\ref{??} as
function of bound cations.
As expected~\cite{seelig87}, the headgroup order parameter decrease due 
to bound cations is related to the orientration of headgroup more parallel
to membrane normal. The headgroup orientational effect is stronger
in Lipid14 parameters than in ECClipid parameters, which is in line
with smaller dipole moment and order parameter changes in ECClipid model.

\subsection{Cation binding affinity in POPC}

@@

%After achieving accurate head group order parameter response with a cationic surfactant, 
%we advanced to quantifying the response to solutions with varying concentrations of salts, namely \ce{NaCl} and \ce{CaCl_2}.
The headgroup order parameter response to aqueous ionic concentration
can be used to measure ion binding affinity in experiments and simulations,
as discussed in section \ref{??} and in previous literature~\cite{seelig87,catte16}.
The headgroup order parameter response of PC lipids to NaCl and CaCl$_2$ concentrations
from experiments (DPPC \cite{akutsu81} and POPC \cite{altenbach84}) and different simulation
models are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
The simulation results of Lipid14 with Aqvist ions \cite{??} taken directly from
Ref. \citenum{catte16} are shown together with the results from Lipid14 simulation
with Dang ions \cite{??} and from ECClipid simulations with ECC-ions.

As already discussed in the recent work by NMRlipids project~\cite{catte16},
the headgroup order parameters are almost unaffected by addition of NaCl
in Lipid14 simulations with Aqvist ion model due a very weak binding of Na$^+$
in PC bilayer. This is in agremeent with experimental data, in contrast
to almost all the other lipid models, which significantly overestimate
Na$^+$ binding in PC bilayers~\cite{catte16}. Despite of good performance
with NaCl, the response of headgroup order parameters to CaCl$_2$ concentration are
significantly overestimated in Lipid14 simulations with Aqvist ion model,
see Fig.~\ref{OrderParameterCHANGESnewMODELS} and \citenum{catte16}.

Wide range of data presented in previous work \cite{catte16} suggested
that CaCl$_2$ binding behaviour cannot be corrected by only improving
ion models. This was not, however, tested for Lipid14 model, so we
ran also Lipid14 simulations with Dang \cite{smith94,chang1999,dang2006} and ECC corrected \cite{??}
ions having more realistic bulk behaviour as Aqvist model \cite{??}.
In line with previous work, the headgroup order parameters response
to CaCl$_2$ concentartion was overestimated also with these ion models,
as seen in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively
\todo{Add OP-response of Lipid14+Dang in Fig. \ref{OrderParameterCHANGESnewMODELS}
  and OP-response of Lipid14+ECC-ions plot in SI}.

Applying ECC correction also for lipids, introduced in section \ref{??},
brings the order parameter response to CaCl$_2$ concentration to a
good agreement with experiments, as shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
This is a significant improvement over all models introduced previously
which in general overstimate the head group order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicates
that the ECC corrected lipid model has a sufficient accuracy for realistic
studies of ion binding details on lipid bilayers.

Ion density profiles along membrane normal coordinate
from different simulations are shown in Fig.~\ref{fig:cacl-dens}.
\todo{This paragraph is to be finalized when we have the surface excess analysis
  done by using Eq.~\ref{surfexcess}.}
This is directly related to the amount of surface excess charge 
as shown along with ion density profiles in  Fig.~\ref{fig:cacl-dens}.
The density profiles from simulations with original Lipid14 \cite{dickson14}
and Dang ions \cite{smith94,chang1999,dang2006} 
show a pronounced peak in the position of the phosphate moieties of POPC. 
The use of a ECC-ion model~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}
or the standard Amber ion model by \AA{qvist}~\cite{aqvist90}
along with Lipid14 does not significantly change it 
(Headgroup order parameter responses for these models in SI).
The new ECC-lipids model with ECC-ions~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}
exhibits on the other hand smaller density in this region 
suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
It is then likely that the models that overestimate head group order parameter response~\cite{catte16}
also overestimate the amount of bound cations.

@@
